Challenging Fuel Cycle Modeling Assumptions: Facility and Time Step Discretization Effects

### 2015-10-20 paper.tex

A fleet reactor archetype was created for Cyclus.  The flexibility of Cyclus
makes it relatively straight forward to create plug-in archetypes to model
anywhere along the spectrum from individual, discrete facilities to
continuous flow fleet facilities.  In order to contrast well with individual,
single-facility granularity modeling, the fleet archetype created here falls more
toward the aggregate, continuous flow side of the spectrum.

@@

The scenario modeled approximates a transition from a the current once-through light water
reactor (LWR) fleet to a full sodium fast reactor (SFR) fuel cycle within evaluation group 23 (EG-23)
in the DOE's Advanced Nuclear Fuel Cycle Options report
\cite{wigeland_nuclear_2014}. Scenario and transition details were patterned
after ongoing work by the DOE in their Fuel Cycle Options campaign.

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.47\textwidth]{exp2/flow.eps}
    \caption[Material flow paths]{Potential material flow paths between facility types.}
    \label{fig:flow}
\end{figure}

A pre-release of Cyclus version 1.4 was used for this analysis.  Archetypes
used were the standard ones provided with Cyclus in the included Cycamore
library - also a pre-release of version 1.4.  A custom storage archetype was
used because Cycamore did not yet provide one.  The standard cycamore reactor archetype
was used for both LWR and SFR reactor types in cases MI and QI.  Correspondingly
the custom fleet-based reactor archetype described earlier was used for both reactor
types in cases MF and QF.  Additional configuration for the reactor types is
shown in Table \ref{tab:reactors} (cycle length includes the refueling outage).

@@

\begin{table}[h]
    \caption{Reactor Facility Parameters}
    \centering
    \begin{tabular}{ |r | c c | }
        \hline                       
            & LWR & SFR \\
        \hline                       
        Lifetime (yr)         & 80 & 80 \\
        Cycle length (months) & 18 & 14 \\
        Batch size (kg)       & 29565 & 7490 \\
        Batches per core      & 3 & 5 \\
        \hline                       
    \end{tabular}
    \captionsetup{justification=centering}
    \label{tab:reactors}
\end{table}

The described scenario is used as closely as possible in each of the four
cases.  However, a few small case-dependent changes to the scenarios relating to
reactor configuration are necessary.  For the cases using quarterly time steps
with individual reactors, cycle and outage times must be adjusted to be a
multiple of 3 months.  In order to make the comparison more valid, the cases
with monthly time steps are also adjusted to have identical cycle and outage
times.  The invariants preserved with respect to reactor behavior between all
four cases are shown in Table \ref{tab:invar}.  Table \ref{tab:reactor-detail} shows
the computed/selected configuration for both LWR and SFR reactor types for all
four cases.

@@

\begin{table*}
    \centering
    \captionsetup{justification=centering}
    \caption{Reactor Parameters by Case}
    \begin{tabular}{ |r | c c | c c | }
        \hline                       
                                          & \multicolumn{2}{c |}{LWR}       & \multicolumn{2}{c |}{SFR} \\
                                          & Cases MI,QI & Cases MF,QF (fleet) & Cases MI,QI & Cases MF,QF (fleet)  \\
        \hline                       
        Cycle length (months)             & 15        & 18                & 12        & 15 \\
        Refueling outage (months)         & 3         & 0                 & 3         & 0 \\
        Batch size ($\text{kg} \cdot \text{HM}$) & 29,565    & 29,565            & 8,025     & 40,125 \\
        Power capacity (MWe)              & 1,080     & 900               & 450       & 360 \\
        \hline                       
    \end{tabular}
    \label{tab:reactor-detail}
\end{table*}

In each of the four cases, the initial 100 LWRs are modified to have a zero
length refueling outage with an explicit 900 MWe net power output capacity.  Because the
individual reactor model doesn't have any way to start a reactor mid-cycle,
this was done to avoid having all initial LWRs refuel at the same time.  The
parameters for each case were also selected to keep capacity factors for
individual reactors very roughly around 90\%.

@@

One difference between the fleet and individual facility modeling is the
\emph{quantized shutdown effect}.  If a fleet-based reactor modeling 3-batch
cores is short one half batch of fuel, it will produce $\frac{2.5}{3}\cdot C$
of power for that time step where $C$ is the power capacity of a single
reactor of the fleet (e.g. 1000 MWe).  If an individually modeled reactor is
short a half batch of fuel, then it will produce no power ($0 \cdot C$) that
time step.  Time step duration can also have this effect.  For both fleet and
individual reactor modeling, power goes offline in full time step increments.
For example, if a 3-batch fleet based reactor is short $\frac{1}{2}$ batch of
fuel with a 3 month time step, then it will produce $\frac{2.5}{3} \cdot 3
\text{mo} \cdot C $ of energy.  In contrast, a monthly time step might allow
the same reactor to produce $[\frac{5}{6} \cdot 1 \text{mo} + 1 \cdot 2
\text{mo}] \cdot C $ of energy if the fuel shortage actually ended after only
1 month.  Quantifying this effect directly is difficult.  One way of observing
it is to compare fleet reactor scenarios that have different time step
durations (i.e. cases MF and QF).  Because the fleet reactors have perfect fuel
sharing (described below) and no noise from refueling outages, differences observed will
likely be due to quantized shutdown.

@@

\begin{equation}
    N_{wasted}(t) = \sum\limits_{r \in R_t} \left(H[t-S_{sched}(t,r)] \cdot [1-O(t,r)]  \sum\limits_{t'=F_{prev}(t,r)}^{t} N_b(t',r) \right)
    \label{eq:badshare}
\end{equation}

@@

Quantifying imperfect fuel sharing is a bit difficult, but Figure
\ref{fig:badshare} provides one way to see the \emph{fuel sharing effect}
showing wasted batch-months computed using Equation \ref{eq:badshare}.  Fleet
reactor modeling by design exhibits perfect fuel sharing by design (zero
inefficiency) and so is not included in this figure.  Every batch that is
given to a reactor that ends up not being able to start its cycle at the
scheduled time (due to fuel shortage) adds one to this cumulative for each
time step the cycle start is delayed.

>>>>

A fleet reactor archetype was created for Cyclus.  Although the modeling of
individual, discrete facilities was a motivation for Cyclus, its flexibility
makes it relatively straightforward to create plug-in archetypes to model
anywhere along the spectrum from individual, discrete facilities to
continuous flow fleet facilities.  In order to contrast well with the individual,
single-facility granularity modeling that is part of the standard Cyclus archetypes, 
a fleet archetype was created for this work with aggregate facilitiy behavior and
continuous flow.

@@

The chosen scenario approximates a transition from a the current once-through light water
reactor (LWR) fleet to a full sodium fast reactor (SFR) fuel cycle within evaluation group 23 (EG-23)
in the DOE's Advanced Nuclear Fuel Cycle Options report
\cite{wigeland_nuclear_2014}. Scenario and transition details were patterned
after ongoing work by the DOE in their Fuel Cycle Options campaign.
%% Not sure if we should make such an explicit reference to the ongoing FCO
%% transition work until they publish something. On one hand, it reveals
%% things they are secretive about. On the other hand, we should give them due
%% credit.

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.47\textwidth]{exp2/flow.eps}
    \caption[Material flow paths]{Potential material flow paths between facility types.}
    %% consider a Cyclist-derived visualization here?
    \label{fig:flow}
\end{figure}

%% is it worth waiting until the 1.4 release?  Citing an ambiguous pre-release makes
%% this a little harder to reproduce.
A pre-release of Cyclus version 1.4 was used for this analysis, with the
standard archetypes provided with Cyclus in the included Cycamore
library - also a pre-release of version 1.4.  A custom storage archetype was
developed because Cycamore did not yet provide one.  The standard Cycamore reactor archetype
was used for both LWR and SFR reactor types in cases MI and QI.  Correspondingly
the custom fleet-based reactor archetype described earlier was used for both reactor
types in cases MF and QF.  Additional configuration for the reactor types is
shown in Table \ref{tab:reactors} (cycle length includes the refueling outage).

@@

\begin{table}[h]
    \caption{Reactor Facility Parameters}
    \centering
    \begin{tabular}{ |r | c c | }
        \hline                       
            & LWR & SFR \\
        \hline                       
        Lifetime (yr)         & 80 & 80 \\
        Cycle length (months) & 18 & 14 \\  %% why list this SFR cycle length if it's never used?
        Batch size (kg)       & 29565 & 7490 \\
        Batches per core      & 3 & 5 \\
        \hline                       
    \end{tabular}
    \captionsetup{justification=centering}
    \label{tab:reactors}
\end{table}

The described scenario is used as closely as possible in each of the four
cases.  However, a few small case-dependent changes to the scenarios relating to
reactor configuration are necessary.  For case \emph{QI}, 
cycle and outage times must be adjusted to be a
multiple of 3 months. 
%% this tweak is a distraction.  Why not simply declare the SFR cycle length &
%% outage as 12/3 from the beginning.  Is this because the fuel recipes you
%% are using assume the other parameters and you don't want to sweep that
%% under the rug?
 In order to make the comparison more valid, the cases
with monthly time steps are also adjusted to have identical cycle and outage
times.  The invariants preserved with respect to reactor behavior between all
four cases are shown in Table \ref{tab:invar}.  Table \ref{tab:reactor-detail} shows
the computed/selected configuration for both LWR and SFR reactor types for all
four cases.

@@

\begin{table*}
    \centering
    \captionsetup{justification=centering}
    \caption{Reactor Parameters by Case}
    \begin{tabular}{ |r | c c | c c | }
        \hline                       
                                          & \multicolumn{2}{c |}{LWR}       & \multicolumn{2}{c |}{SFR} \\
                                          & Cases MI,QI & Cases MF,QF (fleet) & Cases MI,QI & Cases MF,QF (fleet)  \\
        \hline                       
        Cycle length (months)             & 15        & 18                & 12        & 15 \\
        Refueling outage (months)         & 3         & 0                 & 3         & 0 \\
        Batch size ($\text{kg} \cdot \text{HM}$) & 29,565    & 88,695            & 8,025     & 40,125 \\
        Power capacity (MWe)              & 1,080     & 900               & 450       & 360 \\
        \hline                       
    \end{tabular}
    \label{tab:reactor-detail}
\end{table*}

In each of the four cases, the initial 100 LWRs are modified to have a zero
length refueling outage with an explicit 900 MWe net power output capacity.  Because the
individual reactor model doesn't have any way to start a reactor mid-cycle,
this was done to avoid having all initial LWRs refuel at the same time.  The
parameters for each case were also selected to keep capacity factors for
individual reactors very roughly around 80\%.

@@

One difference between the fleet and individual facility modeling is the
\emph{quantized shutdown effect}.  If a fleet-based reactor modeling 3-batch
cores is short one half batch of fuel, it will produce $\frac{2.5}{3}\cdot P_r$
of power for that time step where $P_r$ is the power capacity of a single
reactor of the fleet (\emph{e.g.}~1000~MWe).  If an individually modeled reactor is
short a half batch of fuel, then it will produce no power ($0 \cdot P_r$) that
time step.  Time step duration can amplify this effect.  For both fleet and
individual reactor modeling, power goes offline in full time step increments.
Consider an example of a 3-batch reactor with a fuel shortage of $\frac{1}{2}$ batch of
fuel that would be resolve in 1 month.  Depending on the modeling choice, each reactor
would produce the following amounts of energy over a 3 month period:

\begin{itemize}
  \item case \emph{QF}: $\frac{2.5}{3} \cdot 3 \text{mo} \cdot P_r = 2.5P_r$
  \item case \emph{MF}: $[\frac{2.5}{3} \cdot 1 \text{mo} + 1 \cdot 2
    \text{mo}] \cdot P_r = 2.83P_r$
  \item case \emph{QI}: $0 \cdot 3 \text{mo} \cdot P_r = 0$
  \item case \emph{MI}: $[0\cdot 1 \text{mo} + 1 \cdot 2\text{mo}] \cdot P_r = 2 P_r$
\end{itemize}

Quantifying this effect directly is difficult.  One way of observing
it is to compare fleet reactor scenarios that have different time step
durations (i.e. cases MF and QF).  Because the fleet reactors have perfect fuel
sharing (described below) and no noise from refueling outages, differences observed will
likely be due to quantized shutdown.

@@

\begin{equation}
    N_{wasted}(t) = \sum\limits_{r \in R_t} \left(H[t-S_{sched}(t,r)] \cdot [1-O(t,r)]  \sum\limits_{t'=F_{prev}(t,r)}^{t} N_b(t',r) \right)
    \label{eq:badshare}
\end{equation}
%% What is the value of this function during a normal refueling outage,
%% immediately before S_(sched)(t)?  It seems you need another Heavyside
%% function to drop this back down to an expected value of 0 again.
%%
%%          ___________________
%%          |                 |
%%  ________|                 |_____________

@@

Quantifying imperfect fuel sharing is a bit difficult, but Figure
\ref{fig:badshare} provides one way to see the \emph{fuel sharing effect}
showing wasted batch-months computed using Equation \ref{eq:badshare}.  Fleet
reactor modeling, by design, exhibits perfect fuel sharing by design (zero
inefficiency) and so is not included in this figure.  Every batch that is
given to a reactor that ends up not being able to start its cycle at the
scheduled time (due to fuel shortage) adds one to this cumulative for each
time step the cycle start is delayed.

%% What about a different measure of fuel sharing effect?  Perfect sharing
%% in a fleet reactor results in N_o reactors operating.  For individual
%% reactors, that would mean floor(N_o) reactors.  Can't we measure the
%% actual number of reactors that are operating, N_actual, and accumulate
%% floor(N_o) - N_actual as a measure of the effect of fuel sharing?  Would
%% this be substantially different?
